---
title: "Background"
description: "Characterizing Halifax County"
weight: 1
draft: false
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 15, fig.height = 10)

library(dplyr)
library(leaflet)
library(leaflet.extras)
library(leaflet.mapboxgl)
library(tidycensus)
library(tigris)
library(tidyr)
library(here)
library(sf)
library(glue)

token <- Sys.getenv("MAPBOX_TOKEN")
options(mapbox.accessToken = token)
```

```{r, results = "hide"}
create_map <- function(.data, # spatial dataset to use
                       variables, # character vector of names to be used, does not include _estimate or _moe at end. Example: estimate_percent_below_poverty_level_population_for_whom_poverty_status_is_determined
                       group_names, # The names to appear on the radio button widget, determing which variable to display
                       legend_name, # name of the legend
                       label_name, # What precedes the variable value on the hoverover label
                       scale_domain, # over what values is the scale defined? Values outside will be color NA
                       scale_breaks, # what divides the colors of the scale? For a 4 color scale, and example would be c(0, 25, 50, 75, 100) Note endpoints are included
                       unstable_threshold # How many times larger does the estimate have to be than the margin of error to be considered non-null?
) {
  color_scale <- colorBin("BuPu", scale_domain, scale_breaks)
  
  check_unstable <- function(variable) {
    ifelse((.data[[glue("{variable}_estimate")]]) < unstable_threshold * .data[[glue("{variable}_moe")]],
           NA,
           .data[[glue("{variable}_estimate")]])
  }
  
  add_poly_layer <- function(map, variable, group_name, color_scale) {
    addPolygons(map, color = "#444444", weight = 0.5, smoothFactor = 0.5,
                opacity = 1.0, fillOpacity = 0.7,
                fillColor = color_scale(check_unstable(variable)),
                group = group_name,
                label = ~purrr::map(glue("<strong>{NAME.x} County<br/></strong>
                                  <strong>{label_name}</strong>: {.data[[paste0(variable, \"_estimate\")]]}<br/>
                                  <strong>MOE</strong>: {.data[[paste0(variable, \"_moe\")]]}"), htmltools::HTML))
  }
  
  map <- leaflet(.data, width = "100%") %>%
    addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
    addMapPane("border", zIndex = 410)
  
  for (i in 1:length(variables)) {
    map <- map %>%
      add_poly_layer(variable = variables[i], group_name = group_names[i], color_scale = color_scale)
  }
  
  map <- map %>%
    addLegend("bottomright", pal = color_scale, values = .data[[glue("{variables[1]}_estimate")]],
              title = legend_name,
              opacity = .9
    )
  
  if(!is.null(group_names)) {
    map <- map %>%
      addLayersControl(
        baseGroups = group_names,
        options = layersControlOptions(collapsed = FALSE)
      )
  }
  
  map
}
```

```{r, results = "hide"}
age_sex_race_county <- st_read(here::here("data", "original", "ACS", "acs_age_sex_race_county.geojson"))
acs_median_income_county <- st_read(here::here("data", "original", "ACS", "acs_median_income_county.geojson"))
halifax_decennial_data <- st_read(here::here("src", "Data_Ingestion", "halifax_decennial_data.geojson"))

halifax_border <- tigris::counties(state = "VA", cb = TRUE, resolution = "20m") %>% 
  st_as_sf() %>% 
  st_transform(crs = 4326) %>%
  filter(GEOID == 51083)
```

Halifax County is a large, predominantly rural county in Southside Virginia. Its largest town, South Boston, recorded a population just shy of 8,000 during the decennial census, and the county as a whole has seen a recent decline in population from around 36,000 to 34,000 residents in the past decade[^CENSUS]. Black individuals constitute approximately 34% of the county's population and White individuals 60%. Because these two racial groups constitute the vast majority of the county's population, we restrict our focus to them when comparing across race throughout this website. This makeup is more balanced than in many other Virginia counties.

```{r, fig.height = 4}
#BAMMtools::getJenksBreaks(age_sex_race_county$percent_estimate_race_total_population_one_race_black_or_african_american_estimate, 6)
pal <- colorBin("BuPu", domain = c(0, 1), bins = c(0, 10, 22, 35, 50, 75))

create_map(age_sex_race_county, c("percent_estimate_race_total_population_one_race_black_or_african_american", "percent_estimate_race_total_population_one_race_white"),
           group_names = c("Black" ,"White"),
           legend_name = "Percentage Selected Race",
           label_name = "Percentage",
           scale_domain = c(0, 1),
           scale_breaks = c(0, 10, 25, 40, 60, 75, 90, 100),
           unstable_threshold = 1.5) %>%
  addPolylines(data = halifax_border,
               color = "#FC4444",
               opacity = 0.9,
               weight = 2,
               options = pathOptions(pane = "border"))
```

<br>
In keeping with the typical pattern of much of Southside Virginia, Halifax county's population is slightly older than the state as a whole, with a median age of 46.1 years.
<br>

```{r, fig.height = 4}
#BAMMtools::getJenksBreaks(age_sex_race_county$estimate_sex_and_age_total_population_median_age_.years._estimate, 7)
pal <- colorBin("BuPu", domain = c(0, 1), bins = c(20, 28, 36, 41, 45, 50, 60))

leaflet(width = "100%") %>%
  addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
  addPolygons(data = age_sex_race_county,
              fillColor = ~pal(estimate_sex_and_age_total_population_median_age_.years._estimate),
              fillOpacity = 0.7,
              color = "#444444",
              weight = 1,
              label = ~purrr::map(glue("<strong>{NAME.x} County<br/></strong> <strong>Age</strong>: {age_sex_race_county$estimate_sex_and_age_total_population_median_age_.years._estimate} years<br/><strong>MOE</strong>: {age_sex_race_county$estimate_sex_and_age_total_population_median_age_.years._moe}"), htmltools::HTML)) %>%
  addLegend("bottomright", title = "Median Age", pal = pal, values = c(20, 60)) %>%
  addPolylines(data = halifax_border,
               color = "#FC4444",
               opacity = 0.9,
               weight = 2)
```

<br>
Some gaps between Black individuals and White individuals in Halifax county begin to reveal themselves when we explore median income. Across Virginia, Black individuals make consistently less than their White counterparts on aggregate. In Halifax county, the same pattern appears to play out, with the White median income roughly \$48,000 compared to the Black median income of approximately \$30,400. Importantly, the disparities in some neighboring counties appear slightly less pronounced. As we consider incarceration trends in the county, these patterns will serve as important context.
<br>

```{r, fig.height = 4}
#BAMMtools::getJenksBreaks(acs_median_income_county$estimate_median_income_.dollars._household_income_by_race_and_hispanic_or_latino_origin_of_householder_households_one_race.._black_or_african_american_estimate, 6)
pal <- colorBin("BuPu", domain = c(0, 1), bins = c(10000, 35000, 50000, 65000, 100000, 150000))

create_map(acs_median_income_county,
           c("estimate_median_income_.dollars._household_income_by_race_and_hispanic_or_latino_origin_of_householder_households_one_race.._black_or_african_american","estimate_median_income_.dollars._household_income_by_race_and_hispanic_or_latino_origin_of_householder_households_one_race.._white"),
           group_names = c("Black Median Income", "White Median Income"),
           legend_name = "Median Income",
           label_name = "Median income",
           scale_domain = c(10000, 150000),
           scale_breaks = c(10000, 35000, 50000, 65000, 100000, 150000),
           unstable_threshold = 1.5) %>%
  addPolylines(data = halifax_border,
               color = "#FC4444",
               opacity = 0.9,
               weight = 2,
               options = pathOptions(pane = "border"))
```

<br>
It is difficult to discern demographic patterns within Halifax county itself because of the small sample sizes within each census tract and block group. In many cases, this makes the resulting estimates from the Census Bureau highly uncertain, and reduces our confidence in the precise values for the variables of interest. For this reason, to get a rough sense of tract-level demographic variation, we returned to the 2010 decennial census, which collects more data than the more frequent American Community Survey (ACS). However, given that the 2010 census is now a full decade out-of-date, these results should be treated with some caution.

The following maps highlight the distribution of race and age across the county, revealing that census tracts west of South Boston are composed of larger Black communities, and that median ages increase in the more southern rural parts of the county.
<br><br>

#### Race Breakdown by Census Tract

```{r, fig.height = 4}
#BAMMtools::getJenksBreaks(halifax_decennial_data$pct_black_race, 4)
pal <- colorBin("BuPu", domain = c(0, 1), bins = c(0.15, 0.25, 0.4, 0.5, 1))

halifax_decennial_data %>%
  st_transform(crs = 4326) %>%
  #limited dragging and set minimum and maximum zoom settings
  leaflet(width = "100%") %>%
  #added base tiles
  addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
  #added layers of chloropleth maps depicting each race
  addPolygons(fillColor = ~pal(pct_black_race),
              fillOpacity = 0.7,
              color = "#444444",
              weight = 0.5) %>%
  addLegend(position = "bottomright", pal = pal, values = c(0,1), 
            title = "Proportion Black")
```

<br>

#### Median Age by Census Tract

```{r, fig.height = 4}
#BAMMtools::getJenksBreaks(halifax_decennial_data$total_median_age, 4)
pal <- colorBin("BuPu", domain = c(0, 1), bins = c(38, 42, 46, 50))

halifax_decennial_data %>%
  st_transform(crs = 4326) %>%
  #limited dragging and set minimum and maximum zoom settings
  leaflet(width = "100%") %>%
  #added base tiles
  addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
  #added layers of chloropleth maps depicting each race
  addPolygons(fillColor = ~pal(total_median_age),
              fillOpacity = 0.7,
              color = "#444444",
              weight = 0.5) %>%
  addLegend(position = "bottomright", pal = pal, values = c(0,1), 
            title = "Median Age")
```

<br>

While we were limited in our ability to get reliable up-to-date demographic information at the sub-county level, these general patterns provide helpful context as we further consider patterns in incarceration in the county.

[^CENSUS]: U.S. Census Bureau (2020). Annual Estimates of the Resident Population for Counties in Virginia: April 1, 2010 to July 1, 2019 (CO-EST2019-ANNRES-51). Retrieved from https://www.census.gov/data/tables/time-series/demo/popest/2010s-counties-total.html.
