---
title: "Family Structure"
description: "How stable are family units in Halifax?"
tags: ["R", "Visualization"]
weight: 6
draft: false
output: html_document
---

```{r setup, include=FALSE}
#set up r function on website
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 20, fig.height = 10)

library(here)
library(dplyr)
library(gghighlight)
library(tidyr)
library(leaflet)
library(sf)
library(leaflet.extras)
library(leaflet.mapboxgl)

options(mapbox.accessToken = Sys.getenv("MAPBOX_TOKEN"))

theme_set(theme_minimal() + 
            theme(plot.title = element_text(hjust = 0.5, color = "gray10", size = 22), 
                  plot.subtitle = element_text(hjust = 0.5, color = "gray30", face = "italic", size = 18),
                  axis.title = element_text(size = 18, color = "gray10"),
                  axis.text = element_text(size = 16, color = "gray30"),
                  strip.text = element_text(size = 20, color = "gray30"),
                  panel.spacing = unit(4, "lines"),
                  legend.key.size = unit(3, "line"),
                  legend.text = element_text(size = 14, color = "gray30"),
                  legend.title = element_text(size = 20, color = "gray10")))

#read in Teen Birth Data
Teen_Births_VA <- st_read(here::here("src", "Data_Ingestion", "Teen_Births_VA.geojson"))
```

<<<<<<< HEAD
=======
```{r}
halifax_decennial_data <- st_read(here::here("src", "Data_Ingestion", "halifax_decennial_data.geojson"))

BAMMtools::getJenksBreaks(halifax_decennial_data$pct_female_no_husband_household, 4)
pal <- colorBin("BuPu", domain = c(0, 1), bins = c(0, 0.1, 0.15, 0.25, 1))

halifax_decennial_data %>%
  #limited dragging and set minimum and maximum zoom settings
leaflet(width = "100%") %>%
  #added base tiles
  addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
  #added layers of chloropleth maps depicting each race
  addPolygons(fillColor = ~pal(pct_female_no_husband_household),
              fillOpacity = 0.8,
              color = "#444444",
              weight = 0.5) %>%
  addLegend(position = "bottomright", pal = pal, values = c(0,1), 
            title = "Proportion Single Women Households")

```

There were many government data sources used for this research, most prominent of which were the Center for Disease and Prevention’s (CDC) and Department of Health and Human Services (HHS)’ many articles regarding teen births. Other sources include the County Health Rankings website, Youth.gov website, Guttmacher Institute, Child Trends website, and the National Center for Biotechnology Information (NCBI). Many of these sources had information up to 2018, which is still relevant for statistics. However, social and economic consequences of teen pregnancy varied by source which made it difficult to compile an accurate list of consequences. Further research should be done regarding the economic costs of teen pregnancy. 

>>>>>>> 4b9607bf64fc2b9a2f54913164249c9ea985b8b1
### Background

Decades of research has found that family structure has the power to impact a child’s development. For example, children growing up in two parent homes tend to have better health and social outcomes than those who grow up in single parent homes (The Impact of Family Structure on the Health of Children: Effects of Divorce, 2020). Teen pregnancy is generally defined as pregnancy between the ages of 15 and 19 years old. Teen pregnancies are estimated to cost U.S. taxpayers between $9 and $11 billion dollars per year and Virginia is estimated to spend $183 million dollars on factors related to teen pregnancy (Teen Childbearing Is Costly to Taxpayers, 2020).

### Main Findings 

In 2017, the rate of teen pregnancy (per 1,000 women) was 18.8 in the United States. In the same time period, the rate of teen pregnancy (per 1,000 women) in Virginia was 15 and in Halifax county the rate was estimated at 30.8. This put Halifax at the 24th highest rate of teen births out of 134 counties in Virginia. While the number of teen births has been declining since 2003, the high rate led us to investigate factors that could be related to the high rate of teen births in Halifax.  

We found several factors in Halifax that have been linked to higher rates of teen births. For example, Halifax is a rural county with a high rate of poverty, which has been linked to higher teen birth rates (Products - Data Briefs - Number 264 - November 2016, 2016). Additionally, Halifax county only has one publicly funded clinic and zero federally qualified health centers to aid in female contraception usage. By the age of 18, a woman is more than 3 times likely to have a teen birth if they do not use a proper method of contraception. Thus, the small number of clinics may be contributing to the high rate of teen births in Halifax.


```{r}
Teen_B_colors <- colorBin("YlGnBu", domain = c(0, 100), 
                              bins = c(0, 15, 30, 45, 60, 75, 90, 100))
#teen births rate map over 12 years
Teen_Births_VA %>%
  leaflet(options = leafletOptions(minZoom = 7, maxZoom = 12)) %>%
  setView(-78.6569, 38, 7) %>%
  addTiles()%>%
  addResetMapButton() %>%
  addPolygons(fillColor = ~Teen_B_colors(BirthRate2018), color = "black", 
              fillOpacity = 1, group = "2018", opacity = 1, weight = 2,
              highlight = highlightOptions(color = "white",  
                                           bringToFront = TRUE),
              label = ~paste0("Birth Rate: ", (BirthRate2018))) %>%
  addPolygons(fillColor = ~Teen_B_colors(BirthRate2015), color = "black",
              fillOpacity = 1, group = "2015", weight = 2, opacity = 1,
              highlight = highlightOptions(color = "white",
                                           bringToFront = TRUE),
              label = ~paste0("Birth Rate: ", (BirthRate2015))) %>%
  addPolygons(fillColor = ~Teen_B_colors(BirthRate2012), color = "black",
              fillOpacity = 1, group = "2012", weight = 2, opacity = 1,
              highlight = highlightOptions(color = "white", 
                                           bringToFront = TRUE),
              label = ~paste0("Birth Rate: ", (BirthRate2012))) %>%
  addPolygons(fillColor = ~Teen_B_colors(BirthRate2009), color = "black",
              fillOpacity = 1, group = "2009", weight = 2, opacity = 1,
              highlight = highlightOptions(color = "white",
                                           bringToFront = TRUE),
              label = ~paste0("Birth Rate: ", (BirthRate2009))) %>%
  addPolygons(fillColor = ~Teen_B_colors(BirthRate2006), color = "black",
              fillOpacity = 1, group = "2006", weight = 2, opacity = 1,
              highlight = highlightOptions(color = "white",
                                           bringToFront = TRUE),
              label = ~paste0("Birth Rate: ", (BirthRate2006))) %>%
  addLayersControl(baseGroups = c("2018", "2015", "2012",
                                  "2009", "2006"),
                   position = "topleft") %>%
  addLegend(position = "bottomleft", pal = Teen_B_colors, values = c(0,100), 
            title = "Birth Rate") 
```

### Conclusion and Policy Implications 

Teen pregnancy can have drastic effects on the individual, interpersonal, and the school/work environments of both the mother and their child, as well as costly effects on society. As a result of their pregnancy, teen mothers are at elevated risk for low academic achievement, dropping out of high school, being unemployed, using public assistance, and being in poverty than non-pregnant women of the same age (About Teen Pregnancy | CDC, n.d.). Children of teen mothers are also at elevated risk for low academic achievement, low health and behavioral outcomes, high unemployment, and delinquency compared to children of non-teen mothers. Children of teenage mothers are also more likely to become teen parents. Thus, it is will be important to continue to investigate how to mitigate teen births.  

### References 

About Teen Pregnancy | CDC. (n.d.). Center for Disease Control and Prevention. Retrieved July 22, 2020, from https://www.cdc.gov/teenpregnancy/about/index.htm

Guttmacher Data Center. (n.d.). Guttmacher Institute. Retrieved July 25, 2020, from https://data.guttmacher.org/counties/

National Center for Health Statistics. (n.d.). NCHS Data Visualization Gallery - Teen Birth Rates for Age Group 15-19 in the United States by County. Center for Disease Control and Prevention. Retrieved July 25, 2020, from https://www.cdc.gov/nchs/data-visualization/county-teen-births/

Office of Adolescent Health. (2019, May 30). Trends in Teen Pregnancy and Childbearing. U.S. Department of Health & Human Services. https://www.hhs.gov/ash/oah/adolescent-development/reproductive-health-and-teen-pregnancy/teen-pregnancy-and-childbearing/trends/index.html

Pregnancy Prevention | Youth.gov. (n.d.). Youth.Gov. Retrieved July 25, 2020, from https://youth.gov/youth-topics/pregnancy-prevention

Products - Data Briefs - Number 209 - July 2015. (2015, July). Center for Disease and Prevention. https://www.cdc.gov/nchs/products/databriefs/db209.htm

Products - Data Briefs - Number 264 - November 2016. (2016, November). Center for Disease Control and Prevention. https://www.cdc.gov/nchs/products/databriefs/db264.htm#:%7E:text=From%202007%20through%202015%2C%20the%20rates%20in%20these%20counties%20were,greater%20than%20the%20national%20rate.

Teen births. (n.d.). County Health Rankings. Retrieved July 25, 2020, from https://www.countyhealthrankings.org/explore-health-rankings/measures-data-sources/county-health-rankings-model/health-factors/health-behaviors/sexual-activity/teen-births

Teen Births. (2019, May 24). Child Trends. https://www.childtrends.org/indicators/teen-births

Teen Childbearing is Costly to Taxpayers. (2020). National Conference of State Legislatures. https://www.ncsl.org/research/health/teen-childbearing-is-costly-to-taxpayers.aspx

The impact of family structure on the health of children: Effects of divorce. (2020, November 2). National Institutes of Health. https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4240051/






