---
title: "Substance Abuse"
description: "Comparing Halifax to its neighbors on opioid and alcohol use"
weight: 5
draft: false
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 15, fig.height = 10)
#may not need 
#add packages here
options(knitr.duplicate.label = "allow")
```

### Background 

Approximately three out of four state prisoners and four out of five federal prisoners are detained for alcohol and or drug related offenses[^2]. Additionally, 58% of state prisoners and 63% of sentenced jail inmates meet the criteria for substance abuse disorders (SUDs), compared to 5% of the general adult population[^3]. Not considered in these figures is the high rate of drug-related crime; for example, 17% of state prisoners and 18% of federal inmates committed their offense to obtain money to purchase drugs[^5]. 

Effective substance abuse treatment has been linked with decreased rates of recidivism, but many inmates do not receive sufficient treatment, compounding their difficulties re-integrating into society after release. Unfortunately, 60-80% of drug offenders commit a new drug-related crime upon being released from prison[^1].

Substance abuse intersects with incarceration at recidivism at various levels of the ecological model. Though stable housing is critical for recovery from SUDs, on an individual level, many people struggle to find places to live and become homeless. At the community level, support resources both for substance abuse disorders as well as other related factors, like mental health problems and family stressors may be lacking[^4]. Substance abuse can also have severe impacts on individuals' interpersonal relationships, distancing them from a support network that could otherwise provide important support while re-entering society after release.

On the policy level, strict restrictions on eligibility for employment, housing, and various welfare benefits directly linked to past substance use may present significant obstacles for formerly incarcerated individuals to effectively re-enter society. Systemic racism baked into the criminal justice system also leads to vast racial disparities in substance related arrests, convictions, and felonies.

In the United States, substance use and abuse has been thoroughly entangled with incarceration for decades, and remains a critical component of any comprehensive analysis of incarceration trends.

### Main Findings 

_This paragraph does not seem consistent with the data in the chart?_

_The prevailing opioid epidemic has been felt strongly in rural America. However, data from the Center for Disease Control on opioid usage in Virginia counties suggests a positive perspective on the state of the opioid epidemic in Halifax County[^6]. While prescription rates peaked in 2010, they have been steadily decreasing since. In fact, Halifax seems to fare better than Virginia as a whole; opioid prescription rates in the state have decreased, but at a lesser rate and have peaked more frequently in 2012 and 2015._

The prevailing opioid epidemic has been felt strongly in rural America, and data from the Centers for Disease Control provides mixed messages about the state of the epidemic in Halifax County. While it is encouraging to see prescription rates on a rapid decline, this trend likely tells only part of the story. Given the increased visiblity of the epidemic in recent years, this decline is likely driven more by changes in prescribing practices rather than reduced dependence on the part of those using opioids. In fact, as prescription rates decline, many may turn to heroin or fentanyl, which may be more dangerous to obtain and use, and may also be treated more strictly by the criminal justice system.

```{r}
library(ggplot2)
library(rvest)
library(sf)
library(tigris)
library(leaflet)
library(tidycensus)
library(stringr)
library(dplyr)
library(here)
library(BAMMtools)
library(tidyr)
library(leaflet.mapboxgl)

#set the theme 
theme_set(theme_minimal() +
            theme(plot.title = element_text(hjust = 0.5, color = "gray10", size = 24),
                  plot.subtitle = element_text(hjust = 0.5, color = "gray30", face = "italic", size = 20),
                  axis.title = element_text(size = 20, color = "gray10"),
                  axis.text = element_text(size = 18, color = "gray30"),
                  strip.text = element_text(size = 22, color = "gray30"),
                  panel.spacing = unit(4, "lines"),
                  legend.key.size = unit(3, "line"),
                  legend.text = element_text(size = 16, color = "gray30"),
                  legend.title = element_text(size = 22, color = "gray10")))

options(mapbox.accessToken = Sys.getenv("MAPBOX_TOKEN"))

cbbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r}
#read CSV
datasource <- readr::read_csv(here::here("data",
                                         "original", "Substance_Abuse",
                                         "CDC Opioid Prescription Data - Halifax (1).csv")) %>%
  pivot_longer(cols = c(Halifax, `US Total`))

# total <- datasource$`US Total`

#plot the data
ggplot(data = datasource, aes(x = Year, y = value, color = name)) +
  #plot of Halifax and VA state over time
  geom_line(lwd = 1.1) +
  geom_point(size = 2) +
  labs(y = 'Opioid Prescription Rates per 100 People') +
  #relabel the scale appropriately
  scale_color_manual(name = "Location", labels = c("Halifax, VA", "US"), values = cbbPalette) +
  scale_x_continuous(labels = seq(2006, 2018), breaks = seq(2006, 2018)) +
  theme(panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.position = "bottom")
```
```{r, results = "hide"}
#load VA border geographical data
va_borders <- get_acs(table = "B01003", geography = "county", year = 2018, state = "VA",
                      survey = "acs5", geometry = TRUE, cache_table = TRUE) %>% st_transform(crs = 4326)
```
<br>

A map of the same data provides a spatial context for Halifax’s opioid prescription rates compared to other Virginia counties, as well as the change in prescription rates over time for Virginia. We see that not all rural counties are necessarily at similar risk: counties in southwest Virginia tend to have higher prescription rates than rural counties in Southside Virginia. Note that even the lower ends of the scale exceed 100 prescriptions per 100 people, suggesting that many people are linked to multiple prescriptions and emphasizing the magnitude of the epidemic.

<br>

```{r, fig.height = 8}
#get data from 2017 and 2018 separately bc it has a different format
data_2018_and_2017 <-readr::read_csv(here::here("data",
                                                "original", "Substance_Abuse",
                                                "CDC Opioid Prescription Data - 2018 (1).csv"))

#get the relevant data - merged with va borders data - to create maps for each year
create.map <- function(year) {
  link <- paste(paste("http://www.cdc.gov/drugoverdose/maps/rxcounty", year, sep = ""), ".html", sep = "")
  rate_for_year <- html(link)
  tables <- rate_for_year %>% html_table(fill = TRUE)
  table_for_year <- tables[[1]]
  rate_description <- paste(year, "Prescribing Rate")
  table_for_year <- filter(table_for_year, State == 'VA')
  #merge new_tbl with va_borders data
  new_tbl <- merge(va_borders, table_for_year, by.x = "GEOID", by.y = "FIPS County Code")
  column <- new_tbl[ ,8]
  column$geometry = NULL
  #conver the column containing opiod prescription rates to a numeric column instead of character
  column <- as.numeric(sapply(column, noquote))
  new_tbl <- mutate(new_tbl, 'Rate' = column)
}

#special create_maps function for 2017 and 2018
create.map.special <- function(year, table) {
  table_for_year <- filter(table, Year == year)
  new_tbl <- merge(va_borders, table_for_year, by.x = "GEOID", by.y = "State/County FIPS Code", all.x = TRUE)
  new_tbl <- rename(new_tbl, 'Rate' = 'Opiod Prescription Rate per 100')
}

#create corresponding maps for all the years
mapping_2017 <- create.map.special("2017", data_2018_and_2017)
mapping_2018 <- create.map.special("2018", data_2018_and_2017)
mapping_2006 <- create.map("2006")
mapping_2007 <- create.map("2007")
mapping_2008 <- create.map("2008")
mapping_2009 <- create.map("2009")
mapping_2010 <- create.map("2010")
mapping_2011 <- create.map("2011")
mapping_2012 <- create.map("2012")
mapping_2013 <- create.map("2013")
mapping_2014 <- create.map("2014")
mapping_2015 <- create.map("2015")
mapping_2016 <- create.map("2016")

#combine all possible rates from all years together into one dataset
all_rates <- c(mapping_2006$Rate, mapping_2007$Rate, mapping_2008$Rate, mapping_2009$Rate,
               mapping_2010$Rate, mapping_2011$Rate, mapping_2012$Rate, mapping_2013$Rate,
               mapping_2014$Rate, mapping_2015$Rate, mapping_2016$Rate, mapping_2017$Rate,
               mapping_2018$Rate)
all_rates <- na.omit(all_rates)

#use Jenks breaks to see ideal disribution of Opioid Prescription Rates for map scale
generic_bins <-getJenksBreaks(all_rates, k = 6)
generic_bins <- sapply(generic_bins, round)
generic_palette <-colorBin("BuPu", domain = all_rates, bins = c(0, 60, 120, 200, 360, 580))

#label function for each respective year and map
generate.label <- function(year_mapping) {
  my_label <- paste("<strong>", year_mapping$NAME,"<br/></strong>", "Rate: ", year_mapping$Rate, "<br/>",
                    sep="") %>%
    lapply(htmltools::HTML)
}

#create the leaflet
leaflet(width = "100%") %>%
  addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
  #add polygon for each year with respective template, label, and data
  addPolygons(data = mapping_2016, fillColor = ~generic_palette(mapping_2016$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2016',
              label = generate.label(mapping_2016),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2015, fillColor = ~generic_palette(mapping_2015$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2015',
              label = generate.label(mapping_2015),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2014, fillColor = ~generic_palette(mapping_2014$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2014',
              label = generate.label(mapping_2014),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2013, fillColor = ~generic_palette(mapping_2013$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2013',
              label = generate.label(mapping_2013),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2012, fillColor = ~generic_palette(mapping_2012$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2012',
              label = generate.label(mapping_2012),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2011, fillColor = ~generic_palette(mapping_2011$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2011',
              label = generate.label(mapping_2011),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2010, fillColor = ~generic_palette(mapping_2010$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2010',
              label = generate.label(mapping_2010),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2009, fillColor = ~generic_palette(mapping_2009$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2009',
              label = generate.label(mapping_2009),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2008, fillColor = ~generic_palette(mapping_2008$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2008',
              label = generate.label(mapping_2008),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2007, fillColor = ~generic_palette(mapping_2007$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2007',
              label = generate.label(mapping_2007),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2006, fillColor = ~generic_palette(mapping_2006$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2006',
              label = generate.label(mapping_2006),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2017, fillColor = ~generic_palette(mapping_2017$Rate),
              weight = 1, color =  "#444444", fillOpacity = 0.7, group = '2017',
              label = generate.label(mapping_2017),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addPolygons(data = mapping_2018, fillColor = ~generic_palette(mapping_2018$Rate),
              weight = 1, color = "#444444", fillOpacity = 0.7, group = '2018',
              label = generate.label(mapping_2018),
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addLegend(pal = generic_palette, values = generic_bins,
            title = paste("Opiod Prescription Rate per 100 People"), position = "bottomright") %>%
  #layer for each year
  addLayersControl(baseGroups = c("2006", "2007", "2008", "2009", "2010",
                                  "2011", "2012", "2013", "2014", "2015", "2016", "2017", "2018"),
                   options = list(collapsed = FALSE))
```
<br>

The Virginia Department of Health Office of Epidemiology provides further insight into substance abuse in Halifax county. Data on unintentional drug overdose in Halifax indicate that the majority of overdoses are due to opioids, followed by other types of drugs and with the smallest portion being heroin related[^8]. As noted before, it will be interesting to see if the relative proportion of heroin overdoses continues to increase as prescription rates decline. If so, this could represent an important issue for Halifax to target given both the potential health and criminal consequences involved in heroin use.

<br>
```{r}
#read in data source
datasource <- readr::read_csv(here::here("data",
                                         "original", "Substance_Abuse",
                                         "Ed Data VDH - Copy of Halifax.csv")) %>%
  mutate(Drug = ifelse(Drug == "Opiod", "Opioid", Drug))

# ggplot for the data
ggplot(datasource, aes(x =Year, y =Count, fill=Drug))+
  geom_area(alpha = 0.7) + 
  scale_fill_manual(values = cbbPalette) +
  labs(x = "Year", y = "ED Overdose Visits") +
  ggtitle("Emergency Department Overdose Visits in Halifax, VA") +
  theme(legend.title = element_blank())
```
<br>

Another important aspect of substance abuse is alcohol related abuse and overdose. According to data from the National Center for Health Statistics[^9], the percentage of adults reporting heavy or binge drinking has increased steadily since 2011. This trend parallels that of Virginia as a whole, which consistently has a higher excessive drinking rate than Halifax. 


_The corresponding map of this data illustrates how the majority of counties in Virginia – Halifax included –  become redder over time as rates steadily increase._

<br>

```{r}
#load the data source from the CSV
datasource <- readr::read_csv(here::here("data", "original", "Substance_Abuse",
                                         "Excessive Drinking and Alcohol-Impaired Driving Deaths - Halifax (1).csv"))

#grab the '% Excessive Drinking' column for Halifax
list <- c(14, 14, 14, 13, 13, 10 ,10, 11, 8, 8)
#grab the '% Excessive Drinking' column for VA in general
va <- c(17, 17, 17, 17, 17, 16, 16, 16, 16, 16)
ggplot(data = datasource, aes(x = X1)) +
  #add geom_lines for each location
  geom_line(aes(y = list, colour = "#CC6666"), lwd = 1.1) +
  geom_line(aes(y = va, colour = "#9999CC"), lwd = 1.1) +
  labs(x = 'Year', y = '% Excessive Drinking', title = "Excessive Drinking Rates") + scale_x_continuous(breaks = seq(2011, 2020, 2), lim = c(2011, 2020)) +
  #add relevant scaling for y axis
  scale_y_continuous(breaks = seq(5, 20, 5), lim = c(5, 20))  +
  #relabel the scale appropriately
  scale_color_manual(labels = c("VA", "Halifax, VA"), values = c(cbbPalette[2], cbbPalette[1])) +
  guides(color=guide_legend("Location")) +
  theme(legend.title = element_blank())
```
<br>

Lastly, it is important to understand substance abuse and its relationship with incarceration through the community and policy level of the ecological model. The historical legacy of War on Drugs policies (for instance, the differentiation between crack and powder cocaine) combined with statistical over-policing of black and brown neighborhoods have helped to fuel the racial disparities seen in the criminal justice system. Data from the American Civil Liberties Union (ACLU) indicate the racial disparities in marijuana arrests between Black and White individuals (per 100k people). In Virginia, Black people are 3.4 times more likely to be arrested for marijuana possession than White people, a gap that has steadily widened from 2010 to 2018. In Halifax, Black people are 2.9 times more likely to be arrested for marijuana possession[^7]. While this pattern is indiciative of concerning racial disparities, it is difficult to disentangle these figures from the complex interplay of crime, poverty, and other social determinants of incarceration.

<br>
```{r, fig.height = 8}
datasource <- readr::read_csv(here::here("data",
                                         "original", "Substance_Abuse",
                                         "Racial Disparities in Marijuana Arrests by VA County - 2018 (1).csv"))

va_borders <- get_acs(table = "B01003", geography = "county", year = 2018, state = "VA",
                      survey = "acs5", geometry = TRUE, cache_table = TRUE) %>% st_transform(crs = 4326)
datasource <- merge(va_borders, datasource, by.x = "GEOID", by.y = "FIPS Code")
unstable <- filter(datasource, `Meets Population & Reporting Thresholds` == "No")

general_bins <- getJenksBreaks(datasource$`Times more likely Black person arrested in 2018`, k = 6)
general_bins <- sapply(general_bins, round)

general_palette <- colorBin("BuPu", domain = datasource$`Times more likely Black person arrested in 2018`, bins = general_bins)
label <- paste("County: ", datasource$County,"<br/>", "Rate: ",
               paste(datasource$`Times more likely Black person arrested in 2018`, 'x',
                     sep = ""), "<br/>",
               sep="") %>%
  lapply(htmltools::HTML)

leaflet(width = "100%") %>%
  addMapboxGL(style = "mapbox://styles/mapbox/light-v9") %>%
  addPolygons(data = datasource, 
              fillColor = ~general_palette(datasource$`Times more likely Black person arrested in 2018`),
              weight = 1, color = "#444444", fillOpacity = 0.7,
              label = label,
              labelOptions = labelOptions(style = list("font-weight" = "normal", padding = "3px 8px"),
                                          textsize = "13px", direction = "auto")) %>%
  addLegend(pal = general_palette, values = general_bins,
            title = "Rates of Black vs White arrests <br>for marijuana possession (100k people)", position = "bottomright")
```

### Conclusion and Policy Implications 

**We could potentially have this info, and maybe should've included it. I would focus more on the other social determinants stuff below**
_To better understand the relationship between substance abuse, incarceration, recidivism, particularly in Halifax Virginia, it would be helpful to have additional data sources. Primarily, data on substance abuse related crime in the Halifax County jail, such as demographic information, as well as breakdowns on frequencies for types of drug-related offenses, would be particularly relevant. However, such data poses confidentiality concerns for the individuals within the Halifax county jail, and thus larger ethical concerns._

Additionally, data on substance abuse and SUD-related crime at various levels of the ecological model would be extremely helpful. For example, data on housing outcomes for individuals with SUD-related offenses, such as how many such individuals are homeless or struggling to find stable housing? In that vein, it would be helpful to have information about how receptive and inclusive public housing programs in the county are to formerly incarcerated individuals with drug offenses. Furthermore, data on job outcomes and prospects for individuals with SUD-related offenses would be particularly illuminating, such as how many individuals are unemployed or unable to find a permanent job. Though we have geographic data for substance abuse related treatment and help centers in Halifax County, it would be helpful to have data about the outcomes for individuals who seek out such resources. 

[^1]: 5 Startling Facts About Drug Abuse in US Prisons. (2017, July 10). Retrieved August 1, 2020, from https://www.confirmbiosciences.com/knowledge/blog/5-startling-facts-drug-abuse-us-prisons/
[^2]: Bales, W. D., Slyke, S. V., & Blomberg, T. G. (2006). Substance Abuse Treatment in Prison and Community Reentry: Breaking the Cycle of Drugs, Crime, Incarceration, and Recidivism. Georgetown Journal on Poverty Law & Policy, 8(2).
[^3]: Bronson, J., Stroop, J., Zimmer, S., & Berzofsky, M. (2017, June). Drug Use, Dependence, and Abuse Among State Prisoners and Jail Inmates, 2007-2009 (Rep.). Retrieved https://www.bjs.gov/content/pub/pdf/dudaspji0709.pdf
[^4]: Drug Misuse and Addiction. (2020, July 13). Retrieved from https://www.drugabuse.gov/publications/drugs-brains-behavior-science-addiction/drug-misuse-addiction
[^5]: Drug use and crime. (n.d.). Retrieved August 1, 2020, from https://www.bjs.gov/content/dcf/duc.cfm
[^6]: U.S. Opioid Prescribing Rate Maps. (2020, March 05). Retrieved from https://www.cdc.gov/drugoverdose/maps/rxrate-maps.html
[^7]: Virginia. (n.d.). Retrieved from https://graphics.aclu.org/marijuana-arrest-report/VA
[^8]: Virginia Department of Health. (n.d.). Retrieved from https://www.vdh.virginia.gov/opioid-data/emergency-department/
[^9]: Virginia Rankings Data. (n.d.). Retrieved from https://www.countyhealthrankings.org/app/virginia/2020/downloads
